Supporting Information:\\Evidence for black-swan events in animal populations

### 2015-07-21 text/supp.tex
Add missing text in caption
Bankruptcy commit
Major restructuring of text

%!TEX root = anderson-etal-blackswan-timeseries.tex

\begin{centering}
\LARGE
Supporting Information\\[1.0em]
\end{centering}

@@

\section{Details on the heavy-tailed Gompertz probability model}

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors})
were:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align*}
Our prior on $b$ was uninformative between values of $-1$ and $2$. We would
not expect values of $b$ with levels of density dependence as low as $-1$
(very strong inverse density dependence), nor would we generally expect values
above $1$. We allowed values of $b$ above $1$ to allow for non-stationary time
series of growth rates. The estimates of $b$ were well within these bounds.
Our prior on $\lambda$ was very weakly informative within the range of
expected values for population growth and is similar to the default priors
suggested by \citet{gelman2008d} for intercepts of regression models. Our
Half-Cauchy prior on $\sigma$ follows \citet{gelman2006c} and
\citet{gelman2008d} and the specific scale parameter of $2.5$ is based on our
expected range of the value in nature from previous studies
\citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to an
uninformative uniform prior on $\sigma$, but the weakly informative
Half-Cauchy prior substantially sped up chain convergence.

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
exponential rate parameter of $0.1$. We chose a less informative rate
parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
the variance of the t distribution is undefined. This prior gives only a
$7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
to avoid wandering off towards infinity---above approximately $\nu = 20$ the t
distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be informative about the precise value of $\nu$. In the scenario
where the data are uninformative about heavy tails
(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
tailed.

We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential
scale reduction factor---a measure of chain convergence) was greater than
$1.05$ for any parameter or the minimum effective sample size,
$n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples)
for any parameter was less than $200$, we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$
and effective sample size $> 100$ for reliable point estimates and confidence
intervals. In the majority of cases our minimum thresholds were greatly
exceeded. We continued this procedure up to $8000$ iterations ($16000$ total
samples) by which all chains were deemed to have sufficiently converged. These
chain lengths may seem low to those familiar with software such as WinBUGS or
JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan
generally requires far fewer iterations to obtain equivalent effective sample
sizes \citep{stan-manual2014}.

\section{Alternative priors}

To test if the prior on $\nu$ influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on $\nu$ of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}). Note that
the base and weaker priors are relatively flat within the region of $\nu <
20$, which is the region we are mostly concerned about when categorizing populations
as heavy- or thin-tailed.

Our results show that these weaker and stronger priors would have little
influence on our conclusions about heavy-tailed dynamics
(Fig.~\ref{fig:alt-priors}). When the data are informative about tail
behaviour (i.e.\ when there is strong evidence of low $\nu$ values,
upper-right of Fig.~\ref{fig:alt-priors}), the prior has little impact on the
estimate of $\nu$. When the data are less informative about $\nu$ (i.e.\ when
there are no or few tail events and time series are short or noisy), the prior
can pull the estimate of $\nu$ towards larger or smaller values
(Fig.~\ref{fig:alt-priors}). The vast majority of the populations with Pr$(\nu
< 10)$ in the base prior were not altered qualitatively by this range of prior
strength.

@@

We fit four alternative population models to the time-series data to check how
they would influence our conclusions. Our alternative models allowed for
autocorrelation in the residuals, assumed no density dependence, allowed for
observation error, or assumed a Ricker-logistic functional form. The range of
percentages of black swans by taxonomic class cited in the abstract and
results are based on lower and upper limits across our main Gompertz model and
these four alternative models.

@@

@@

We found that we could consistently and precisely recover median posterior
estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
(Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could still
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~\ref{fig:sim-nu} lower panels).

@@

%(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and
%green symbols in the top rows). Likewise, the other Gompertz parameters were
%estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and
%\ref{fig:sim-gompertz-boxplots}, red and green symbols). , overestimate the
%magnitude of process noise, somewhat overestimate $\lambda$, and overestimate
%density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and
%\ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence
%with observation error is a known phenomenon \citep{knape2012}. Fitting a
%model with correctly specified observation error made marginal improvements
%to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and
%\ref{fig:sim-gompertz-boxplots}).

@@

\section{Modelling covariates of heavy-tailed dynamics}

We fit a multilevel beta regression model to the predicted probability of
heavy tails, Pr($\nu < 10$), to investigate potential covariates of
heavy-tailed dynamics. The beta distribution is useful when response data
range on a continuous scale between zero and one \citep{ferrari2004}. We used
a logit link function as is typically used in logistic regression. The model
was as follows:

\begin{align*}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align*}
where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
represents the predicted value for population $i$, class $j$, order $k$, and
species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and
$X_i$ represents a vector of predictors (such as lifespan) for population $i$
with associated $\beta$ coefficients. The intercepts are allowed to vary from
the overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
taxonomic order ($\alpha^\mathrm{order}_k$), and species
($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha
  \; \mathrm{species}}$. Where possible, we also allowed for error
distributions around the predictors by incorporating the standard deviation of
the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\log
\sigma$ around the mean point value as normal distributions (not shown in the
above equation).

We log transformed $\sigma$, time-series length, and lifespan to match the way
they are visually represented in Fig.~\ref{fig:correlates} and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by
two standard deviations to make their coefficients comparable in magnitude
\citep{gelman2008c}. We excluded body length as a covariate because it was
highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$
on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on
the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$
on all other parameters \citep{gelman2006c, gelman2008d}. Compared to normal
priors, the Cauchy priors concentrate more probability density around expected
parameter values while allowing for a higher probability density far into the
tails, thereby allowing the data to dominate the posterior more strongly if it
disagrees with the prior. Our conclusions were not qualitatively changed by
using uniform priors. We fit our models with 5000 total iterations per chain,
2500 warm-up iterations, four chains, and discarding every second sample to
save memory. We checked for chain convergence visually and with the same
criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all
parameters).

To derive taxonomic-order-level estimates of the probability of heavy tails
accounting for time-series length (Fig.~\ref{fig:posteriors}a), we fit a
separate multilevel model with the same structure but with only $\log$
time-series length as a predictor. (In this case, we did not want to control
for intrinsic population characteristics such as density dependence.) Since
our predictors were centered by subtracting their mean value, we obtained
order-level estimates of the probability of heavy tails at mean log
time-series length by adding the posteriors for $\alpha$,
$\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

@@

Many of the silhouette images used in Figs~\ref{fig:nu-coefs},
\ref{fig:correlates} and \ref{fig:posteriors} were obtained from
\texttt{phylopic.org} under Creative Commons licenses. We vectorized the
salmon in Fig.~\ref{fig:nu-coefs} and Fig.~\ref{fig:correlates} ourselves. The
bird in these figures was obtained from \texttt{phylopic.org} under a Creative
Commons Attribution 3.0 Unported license with credit to Jean-Raphaël
Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}). The
silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following
sources (metadata obtained with the help of the rphylopic R package,
\url{https://github.com/sckott/rphylopic}):

@@

\renewcommand{\baselinestretch}{\tighttextstretch}
\normalsize
\bibliographystyle{apalike}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
\renewcommand{\baselinestretch}{\textstretch}
\normalsize

% ------------------------------
% Supplemental Tables
% ------------------------------

\clearpage
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\begin{table}
\begin{footnotesize}

\caption[Summary statistics for the filtered Global Population Dynamics
  Database time series arranged by taxonomic class.]{Summary statistics for the filtered Global Population Dynamics
  Database time series arranged by taxonomic class. Columns are: number of
  populations, number of taxonomic orders, numbers of species, median time
  series length, total number of interpolated time steps, total number of
  substituted zeros, and total number of time steps.}

\smallskip
\begin{tabular}{lrrrrrrrr}
\toprule
\input{../analysis/stat-table.tex}
\label{tab:stats}
\end{tabular}
\end{footnotesize}
\end{table}

@@

\LTcapwidth=\textwidth
\bibpunct{}{}{;}{a}{}{;}

%\singlespacing
\begin{footnotesize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{3.3cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.4cm}>{\RaggedRight}p{1.25cm}}

\caption[All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
  Gompertz population dynamics model.]{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
  Gompertz population dynamics model. Shown are the log abundance time series,
  population descriptions, Global Population Dynamics Database Main IDs,
  citation for the data source or separate verification literature, a
  description of the cause of the black swan events (if known), the
  probability of heavy tails as calculated by our model, and median estimate
  of $\nu$ from our model with 90\% quantile credible intervals indicated in
  parentheses. Red dots on the time series indicate downward black-swan events
  and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were
  explained by a Gompertz model with normally distributed process noise with a
  standard deviation equal to the scale parameter in the fitted t
  distribution.}\\

\toprule
\input{../analysis/cause-table.tex}
\label{tab:causes-supp}
\end{longtable}
\end{footnotesize}
%\onehalfspacing

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{centering}
\clearpage
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-mammals.pdf}\\
Figure~\ref{fig:all-ts} (mammals) continued on next page \ldots

\clearpage
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-birds.pdf}\\
Figure~\ref{fig:all-ts} (birds) continued on next page \ldots

\clearpage
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-insects.pdf}\\
Figure~\ref{fig:all-ts} (insects) continued on next page \ldots

\end{centering}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}

\caption[All filtered time series used in our analysis.]{(fishes, crustaceans,
  gastropods, sharks). All filtered time series used in our analysis. The
  abundances are shown on a log10 vertical axis. Throughout this figure, red
  dots indicate values that were interpolated and blue dots indicate values
  that were recorded as zero but were set to the next lowest observed
  abundance. Numbers before each species name are the GPDD Main ID numbers.}

\label{fig:all-ts}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/priors-gomp-base.pdf}

\caption[Probability density of the Bayesian priors for the Gompertz models.]{
  Probability density of the Bayesian priors for the Gompertz models. From
  left to right and then top to bottom: (1) per capita growth rate at
  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
  log-transformed x axis. Note that the base and weaker priors are relatively
  flat within the region of $\nu < 20$ that we are concerned with. }

\label{fig:priors}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}

\caption[Estimates of $\nu$ from alternative models plotted against the base
Gompertz model estimates of $\nu$.]{Estimates of $\nu$ from alternative models
  plotted against the base Gompertz model estimates of $\nu$. Shown are
  medians of the posterior (dots) and 50\% credible intervals (segments). The
  diagonal line indicates a one-to-one relationship. Different colours
  indicate various taxonomic classes. The grey-shaded regions indicate regions
  of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed
  dynamics. The Gompertz observation error model assumes a fixed standard
  deviation of observation error of $0.2$ on a log scale.}

\label{fig:alt}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-prior-comparison.pdf}

\caption[Estimates of $\nu$ from Gompertz models with alternative priors on
$\nu$.]{Estimates of $\nu$ from Gompertz models with alternative priors on
  $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals
  (segments). The diagonal line indicates a one-to-one relationship. Different
  colours indicate various taxonomic classes. The grey-shaded regions indicate
  regions of disagreement if $\nu = 10$ is taken as a threshold of
  heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are
  illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly
  identical in cases where the data are informative about low values of $\nu$.
  When the data are less informative about low values of $\nu$, the prior can
  slightly pull the estimates of $\nu$ towards higher or lower values.}

\label{fig:alt-priors}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}

\caption[Testing the ability to estimate $\nu$  and the scale parameter of the
process deviations for a given number of samples drawn from a distribution
with a given true $\nu$ value.]{Testing the ability to estimate $\nu$ (top
  panels) and the scale parameter of the process deviations (bottom panels)
  for a given number of samples (columns) drawn from a distribution with a
  given true $\nu$ value (rows). The red lines indicate the true population
  value. When a small number of samples are drawn there may not be samples
  sufficiently far into the tails to recapture the true $\nu$ value; however,
  heavy tails are still distinguished from normal tails in most cases, even
  with only 25 or 50 samples.}

\label{fig:sim-nu}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}

\caption[Simulation testing the Gompertz estimation model when the process
deviation draws were chosen so that $\nu$ could be estimated close to the true
value outside the full population model (``effective $\nu$'' within a CV of
0.2 of specified $\nu$).]{Simulation testing the Gompertz estimation model
  when the process deviation draws were chosen so that $\nu$ could be
  estimated close to the true value outside the full population model
  (``effective $\nu$'' within a CV of 0.2 of specified $\nu$). Upper panels
  show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
  Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation
  runs. We ran the simulations across three population (``true'') $\nu$ values
  (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios:
  (1) 100 time steps and no observation error, (2) 50 time steps and no
  observation error, (3) 50 time steps and observation error drawn from
  $\mathrm{Normal} (0,
  0.2^2)$ but ignored, and (4) 50 time steps with observation error in which
    the quantity of observation error was assumed known. Within each scenario
    the dots represent stochastic draws from the true population distributions
    combined with model fits. Underlayed boxplots show the median,
    interquartile range, and $1.5$ times the interquartile range. }

\label{fig:sim-prob}
\end{center}
\end{figure}

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
residuals and a specified level of observation error. The specific code for
used for the various models in our analysis is available at
\url{https://github.com/seananderson/heavy-tails}.

@@

\renewcommand{\baselinestretch}{\tighttextstretch}
\normalsize
\begin{footnotesize}
\noindent
{\tt
\input{../analysis/mainids}
}
\end{footnotesize}
\renewcommand{\baselinestretch}{\textstretch}
\normalsize
%\end{spacing}

>>>>

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
  lmargin = 2.54cm, rmargin = 2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage{verbatim}
\usepackage[sectionbib]{chapterbib}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabu} % for spacing between rows in longtable
\usepackage{titlesec}
\usepackage{url}
\usepackage{mathtools}
\usepackage{cite}
\usepackage{citesupernumber}

\usepackage{lineno}
\usepackage{xcolor}
\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
\modulolinenumbers[2]

% Linux Libertine:
\usepackage{textcomp}
\usepackage[sb]{libertine}
\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
\usepackage[cal=boondoxo]{mathalfa} % mathcal
\useosf % osf for text, not math
\usepackage[supstfm=libertinesups,%
  supscaled=1.2,%
  raised=-.13em]{superiors}

\mathchardef\mhyphen="2D % math hyphen

\textheight 22.0cm

\usepackage[round,sectionbib]{natbib}
\bibpunct{(}{)}{;}{a}{}{;}

\setlength\parskip{0.10in}
\setlength\parindent{0in}

% Fix line numbering and align environment
% http://phaseportrait.blogspot.ca/2007/08/lineno-and-amsmath-compatibility.html
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

% remove numbers in front of sections:
\makeatletter
\renewcommand\@seccntformat[1]{}
\makeatother

\title{Supporting Information:\\Evidence for black-swan events in animal
  populations}

\author{
Sean C. Anderson$^{1,2}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

\begin{document}

\input{../analysis/values} % R output

\maketitle

\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\textsuperscript{2}School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada

\textsuperscript{3}School of Aquatic and Fishery Sciences, University of
Washington, Box 355020, Seattle, WA 98195, USA

\linenumbers
\onehalfspacing

%\begin{centering}
%\LARGE
%Supporting Information\\[1.0em]
%\end{centering}

@@

@@

We fit four alternative population models to the time-series data to check how
they would influence our conclusions. Our alternative models allowed for
autocorrelation in the residuals, assumed no density dependence, allowed for
observation error, or assumed a Ricker-logistic functional form.

@@

\section{Implementation of skewed Student t distribution in Stan}

The probability density for the skewed Student t distribution described in the
methods section translates to the log probability:

\[
\mathrm{Skewed\,Student\mhyphen t}(x) =
\begin{dcases}
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\]

Since Stan only needs the probability density up to an additive constant, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student's t distribution implemented in
Stan\footnote{See the Stan users mailing list:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

\begin{verbatim}
real skew_student_t_log(real y, real nu, real mu, real sigma, real skew) {
  real lp;
  lp <- log(skew) - log1p(square(skew));
  if (y < mu)
    return lp + student_t_log(y * skew, nu, mu * skew, sigma);
  else
    return lp + student_t_log(y / skew, nu, mu / skew, sigma);
}
\end{verbatim}

@@

In the following image, the red lines indicate the true population value. Dots
represent the median estimate the lines represent the 80\% credible interval.
When a small number of samples are drawn there may not be samples sufficiently
far into the tails to recapture the true $\nu$ value; however, heavy tails are
still distinguished from normal tails in most cases, even with only 25 or 50
samples:

\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\end{center}

We found that we could consistently and precisely recover median posterior
estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
(upper panels). At smaller samples we could still usually qualitatively
distinguish heavy from not-heavy tails, but the model tended to underestimate
how heavy the tails were. At the same time, at smaller sample sizes, the model
tended to overestimate how large the scale parameter was (lower panels).

@@

@@

In the following image, upper panels show the distribution of median
$\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution
of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across
three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\
approximately normal) and four scenarios: (1) 100 time steps and no
observation error, (2) 50 time steps and no observation error, (3) 50 time
steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
ignored, and (4) 50 time steps with observation error in which the quantity of
observation error was assumed known. Within each scenario the dots represent
stochastic draws from the true population distributions combined with model
fits. Underlayed boxplots show the median, interquartile range, and $1.5$
times the interquartile range:

%\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
%\label{fig:sim-prob}
\end{center}
%\end{figure}

@@

Many of the silhouette images used in Figs 2, and 3 were obtained from
\texttt{phylopic.org} under Creative Commons licenses. We vectorized the
salmon in Fig.~2. The bird in these figures was obtained
from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
license with credit to Jean-Raphaël Guillaumin {[}photography{]} and T.
Michael Keesey {[}vectorization{]}). The silhouettes in
Fig.~3 were obtained from the following sources (metadata
obtained with the help of the rphylopic R package,
\url{https://github.com/sckott/rphylopic}):

@@

\bibliographystyle{apalike}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
%
% ------------------------------
% Supplemental Tables
% ------------------------------

%\clearpage
%\renewcommand{\thetable}{S\arabic{table}}
%\setcounter{table}{0}

%\renewcommand{\thefigure}{S\arabic{figure}}
%\setcounter{figure}{0}

% \begin{centering}
% \clearpage
% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-mammals.pdf}\\
% Figure~\ref{fig:all-ts} (mammals) continued on next page \ldots
%
% \clearpage
% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-birds.pdf}\\
% Figure~\ref{fig:all-ts} (birds) continued on next page \ldots
%
% \clearpage
% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-insects.pdf}\\
% Figure~\ref{fig:all-ts} (insects) continued on next page \ldots
%
% \end{centering}

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
%
% \caption[All filtered time series used in our analysis.]{(fishes, crustaceans,
%   gastropods, sharks). All filtered time series used in our analysis. The
%   abundances are shown on a log10 vertical axis. Throughout this figure, red
%   dots indicate values that were interpolated and blue dots indicate values
%   that were recorded as zero but were set to the next lowest observed
%   abundance. Numbers before each species name are the GPDD Main ID numbers.}
%
% \label{fig:all-ts}
% \end{center}
% \end{figure}

%\clearpage

@@

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
residuals and a specified level of observation error. The specific code for
the various models in our analysis is available at
\url{https://github.com/seananderson/heavy-tails}.

@@

\begin{footnotesize}
%\renewcommand{\baselinestretch}{1.11}
\noindent
{\tt
\input{../analysis/mainids}
}
%\renewcommand{\baselinestretch}{\textstretch}
\end{footnotesize}
\normalsize
%\end{spacing}

\end{document}
